Appendix 6: sdmTMB model validation and speed-testing methods

### 2022-01-06 doc/appendix-binomial-inla.Rmd
Appendix edits

---
title: "Appendix 4: Dynamic species distribution model comparison between INLA, inlabru, and sdmTMB"
output:
  pdf_document:
    highlight: tango
    toc: true
    number_sections: true
---

@@

@@

In this example, we will fit a presence-absence species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
We will use a binomial distribution and a logit link.
We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
INLA requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:

```{r}
d <- pcod %>% filter(year >= 2007)
yr_lu <- data.frame(year = unique(d$year), i_year = seq_along(unique(d$year)))
d <- dplyr::left_join(d, yr_lu)
```

In this example, we will use Penalized Complexity (PC) priors on the range and random field standard deviation ("sigma") for models built in INLA and sdmTMB.
The values amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).

@@

In INLA a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `priors = sdmTMBpriors(ar1_rho = normal(0, 1))`).

@@

m_inla <- inla(formula,
  data = inla.stack.data(sdat),
  control.predictor = list(compute = TRUE, A = inla.stack.A(sdat)),
  family = "binomial",
  control.family = list(link = "logit"),
  control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
  control.inla = list(int.strategy = "eb", strategy = "gaussian"),
  control.compute = list(config = TRUE)
)
tictoc::toc() # stop timing

@@

We can then compare the distribution of these sdmTMB simulations to the marginal posterior densities from INLA.
From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the INLA model posterior.

@@

@@

```{r inlabru-pred, out.width="6in"}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu)
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))
ggplot(as.data.frame(pred_bru)) +
  geom_raster(aes(X, Y, fill = mean)) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

>>>>

---
title: "Appendix 4: Dynamic species distribution model comparison between INLA, inlabru, and sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
---

@@

In this example, we will fit a presence-absence species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
We will use a binomial distribution and a logit link.

We will start by loading necessary packages:

@@

We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
INLA requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:

```{r}
d <- pcod %>% filter(year >= 2007)
yr_lu <- data.frame(year = unique(d$year), i_year = seq_along(unique(d$year)))
d <- dplyr::left_join(d, yr_lu, by = "year")
```

We will use Penalized Complexity (PC) priors on the range and random field standard deviation ("sigma") for models built in INLA, inlabru, and sdmTMB.
The values amount to the probability the range is below a lower limit (e.g., `P(range < 5) = 0.05`) and that the sigma is above an upper limit (e.g., `P(sigma > 5) = 0.05`).

@@

In INLA, a PC-prior can be used for the temporal autoregressive parameter, i.e., a prior that defines the probability the autocorrelation parameter is greater than a given value (e.g., `P(cor > 0) = 0.9`), but for now, sdmTMB only takes a Gaussian prior (e.g., `priors = sdmTMBpriors(ar1_rho = normal(0, 1))`).

@@

m_inla <- inla(formula,
  data = inla.stack.data(sdat),
  control.predictor = list(compute = TRUE, A = inla.stack.A(sdat)),
  family = "binomial",
  control.family = list(link = "logit"),
  control.fixed = list(expand.factor.strategy = "inla", mean = 0, prec = 1 / (100 * 100)),
  control.inla = list(int.strategy = "eb", strategy = "gaussian")
)
tictoc::toc() # stop timing

@@

We can then compare the distribution of these sdmTMB simulations to the marginal posterior densities from INLA.
From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.
Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the INLA model posterior assuming we had set `control.compute = list(config = TRUE)`.

@@

In this plot, the blue histograms are from sdmTMB and the density lines are from INLA:

@@

```{r inlabru-pred, out.width="6in"}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))
ggplot(as.data.frame(pred_bru)) +
  geom_raster(aes(X, Y, fill = mean)) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

### 2022-01-06 doc/appendix-model.Rmd
Appendix edits

---
title: "Appendix 1: sdmTMB statistical model description"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: true
fig_caption: true
urlcolor: blue
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
---

@@

* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
<!-- should alpha_g be bold as in the equation above? --><!-- EW: yes -- alpha_g should be bolded here -->
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

### IID spatiotemporal random fields
<-- EW: as above, I changed epsilon to be subscripted with s throughout this section-->
IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as

@@

## IID random or multi-level intercepts
<-- EW: make alpha bold here, as above?-->
$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \alpha_{g} + \ldots \right),\\
  \alpha_g &\sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right),\\
\end{aligned}
$$

@@

$$
\operatorname{Binomial} \left(N, \mu \right)
$$
where $N$, is the size or number of trials, and $\mu$ is the probability of success for each trial.
If $N = 1$, the distribution becomes the Bernoulli distribution.
Internally, the distribution is parameterized as the [robust
version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
<-- EW: Put lme4 and glmmTMB in '' too?-->
Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:

@@

$$
\operatorname{Beta} \left(\mu \phi, (1 - \mu) \phi \right)
$$
<<- EW: betareg in ''?-->
where $\mu$ is the mean and $\phi$ is a precision parameter.
This parametrization follows @ferrari2004 and the betareg R package [@cribari-neto2010].
The variance is $\mu (1 - \mu) / (\phi + 1)$.

@@

$$
\operatorname{Gamma} \left( \phi, \frac{\mu}{\phi}  \right)
$$
<-- EW: check this, I think the Gamma variance should be mu * phi2 -->
where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale. 
The mean is $\mu$ and variance is $\mu \cdot \phi^2$.

@@

TODO: did this get reversed above? Isn't it set up so the mean(x) is mu?
<-- EW: the lognormal definition above includes the bias correction phi2/2, but that's not in the standard implementations in R, etc. I think I used those R functions or referenced Wikipedia in writing the equations out. So we should pick one or the other. I think just changing the definition to not include the bias correction would keep it consistent with R, https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Lognormal.html-->
Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.

>>>>

---
title: "Appendix 1: sdmTMB statistical model description"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
fig_caption: true
urlcolor: blue
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
---

@@

* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

<!-- should alpha_g be bold as in the equation above? -->
<!-- EW: yes -- alpha_g should be bolded here -->

@@

### IID spatiotemporal random fields
<!-- EW: as above, I changed epsilon to be subscripted with s throughout this section-->
IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as

@@

## IID random or multi-level intercepts
<!-- EW: make alpha bold here, as above?-->
$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \alpha_{g} + \ldots \right),\\
  \alpha_g &\sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right),\\
\end{aligned}
$$

@@

$$
\operatorname{Binomial} \left(N, \mu \right)
$$
where $N$, is the size or number of trials, and $\mu$ is the probability of success for each trial.
If $N = 1$, the distribution becomes the Bernoulli distribution.
Internally, the distribution is parameterized as the [robust
version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
<!-- EW: Put lme4 and glmmTMB in '' too?-->
Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:

@@

$$
\operatorname{Beta} \left(\mu \phi, (1 - \mu) \phi \right)
$$
<!-- EW: betareg in ''?-->
where $\mu$ is the mean and $\phi$ is a precision parameter.
This parametrization follows @ferrari2004 and the betareg R package [@cribari-neto2010].
The variance is $\mu (1 - \mu) / (\phi + 1)$.

@@

$$
\operatorname{Gamma} \left( \phi, \frac{\mu}{\phi}  \right)
$$
<!-- EW: check this, I think the Gamma variance should be mu * phi2 -->
where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale. 
The mean is $\mu$ and variance is $\mu \cdot \phi^2$.

@@

TODO: did this get reversed above? Isn't it set up so the mean(x) is mu?
<!-- EW: the lognormal definition above includes the bias correction phi2/2, but that's not in the standard implementations in R, etc. I think I used those R functions or referenced Wikipedia in writing the equations out. So we should pick one or the other. I think just changing the definition to not include the bias correction would keep it consistent with R, https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Lognormal.html-->
Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.

### 2022-01-06 doc/appendix-pcod-VAST-tweedie.Rmd
Appendix edits

---
title: "Appendix 5: example of index standardization of fishery-independent survey data with VAST"
output:
  pdf_document:
    highlight: tango
---

>>>>

---
title: "Appendix 5: Comparison of index standardization of fishery-independent survey data with VAST and sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: false
    number_sections: true
---

### 2022-01-06 doc/appendix-pcod-sdmTMB.Rmd
Appendix edits

---
title: "Appendix 2: Example of index standardization of fishery-independent survey data with sdmTMB"
output:
  pdf_document:
    highlight: tango
---

>>>>

---
title: "Appendix 2: Example of index standardization of fishery-independent survey data with sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
---

### 2022-01-06 doc/appendix-snow-sdmTMB.Rmd
Appendix edits

# Background

@@

# Model validation

@@

<!-- # ```{r} -->
<!-- # library(future) -->
<!-- # plan(multisession) -->
<!-- # m_cv <- sdmTMB_cv(count ~ 1 + nao + (1 | year_f), -->
<!-- #   spatial_varying = ~ 0 + nao, -->
<!-- #   time = "year",  -->
<!-- #   spatial = "on",  -->
<!-- #   spatiotemporal = "iid", -->
<!-- #   family = nbinom2(link = "log"), -->
<!-- #   # reml = TRUE, -->
<!-- #   use_initial_fit = TRUE, -->
<!-- #   data = snow, mesh = mesh, -->
<!-- #   k_folds = 10 -->
<!-- # ) -->
<!-- # m_cv -->
<!-- # ``` -->

>>>>

# Introduction

@@

@@

### 2022-01-06 doc/appendix-validation.Rmd
Appendix edits

---
title: "Appendix 6: sdmTMB model validation and speed-testing methods"
output:
  bookdown::pdf_document2:
    toc: true
number_sections: false
fig_caption: true
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
---

>>>>

---
title: "Appendix 6: sdmTMB model validation and speed-testing methods"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
number_sections: false
fig_caption: true
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
---
